---
title: 'Foundations for statistical inference - Confidence intervals'
author: ""
output:
  html_document:
    
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE, echo = TRUE)
```

If you have access to data on an entire population, say the opinion of every adult in the United States on whether or not they think climate change is affecting their local community, it's straightforward to answer questions like, "What percent of US adults think climate change is affecting their local community?". Similarly, if you had demographic information on the population you could examine how, if at all, this opinion varies among young and old adults and adults with different leanings. If you have access to only a sample of the population, as is often the case, the task becomes more complicated. What is your best guess for this proportion if you only have data from a small sample of adults? This type of situation requires that you use your sample to make inference on what your population looks like.

<div id="boxedtext">
**Setting a seed:** You will take random samples and build sampling distributions in this lab, which means you should set a seed on top of your lab. If this concept is new to you, review the lab on probability.
</div>

## Getting Started

### Load packages

In this lab, we will explore and visualize the data using the **tidyverse** suite of packages, and perform statistical inference using **infer**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(infer)
```

### The data

A 2019 Pew Research report states the following:

To keep our computation simple, we will assume a total population size of 100,000 (even though that's smaller than the population size of all US adults).

> Roughly six-in-ten U.S. adults (62%) say climate change is currently affecting their local community either a great deal or some, according to a new Pew Research Center survey.
>
>**Source:** [Most Americans say climate change impacts their community, but effects vary by region](https://www.pewresearch.org/fact-tank/2019/12/02/most-americans-say-climate-change-impacts-their-community-but-effects-vary-by-region/)

In this lab, you will assume this 62% is a true population proportion and learn about how sample proportions can vary from sample to sample by taking smaller samples from the population. We will first create our population assuming a population size of 100,000. This means 62,000 (62%) of the adult population think climate change impacts their community, and the remaining 38,000 does not think so.

```{r}
us_adults <- tibble(
  climate_change_affects = c(rep("Yes", 62000), rep("No", 38000))
)
```

The name of the data frame is `us_adults` and the name of the variable that contains responses to the question *"Do you think climate change is affecting your local community?"* is `climate_change_affects`.

We can quickly visualize the distribution of these responses using a bar plot.

```{r bar-plot-pop, fig.height=2.5, fig.width=10}
ggplot(us_adults, aes(x = climate_change_affects)) +
  geom_bar() +
  labs(
    x = "", y = "",
    title = "Do you think climate change is affecting your local community?"
  ) +
  coord_flip() 
```

We can also obtain summary statistics to confirm we constructed the data frame correctly.

```{r summ-stat-pop, results = TRUE}
us_adults %>%
  count(climate_change_affects) %>%
  mutate(p = n /sum(n))
```

In this lab, you'll start with a simple random sample of size 60 from the population.

```{r sample}
set.seed(100)
n <- 60
samp <- us_adults %>%
  sample_n(size = n)
```

# Excercise 1.  What percent of the adults in your sample think climate change affects their local community? **Hint:** Just like we did with the population, we can calculate the proportion of those **in this sample** who think climate change affects their local community.

**Insert your answer here**
60  % of the adults in my sample think climate change affects their local community

```{r summ-stat-samp, results = TRUE}
samp %>%
  count(climate_change_affects) %>%
  mutate(p = n /sum(n))
```
# Excercise 2.  Would you expect another student's sample proportion to be identical to yours? Would you expect it to be similar? Why or why not?

**Insert your answer here**
I do not expect my sample proportion to be identical to another student's because sample_n() produces a random sample each time it is called. As a result, different samples may end up including different proportions of the subgroups within the population. Since the sample of 60 is the same for both me and another student and is also relatively small (less than 1% of the population), my proportion will be somehow similar to another student's because each draw does not significantly change the proportion of the two subgroups in the population.

## Confidence intervals

Return for a moment to the question that first motivated this lab: based on this sample, what can you infer about the population? With just one sample, the best estimate of the proportion of US adults who think climate change affects their local community would be the sample proportion, usually denoted as $\hat{p}$ (here we are calling it `p_hat`). That serves as a good **point estimate**, but it would be useful to also communicate how uncertain you are of that estimate. This uncertainty can be quantified using a **confidence interval**.

One way of calculating a confidence interval for a population proportion is based on the Central Limit Theorem, as $\hat{p} \pm z^\star SE_{\hat{p}}$ is, or more precisely, as \[ \hat{p} \pm z^\star \sqrt{ \frac{\hat{p} (1-\hat{p})}{n} } \]

Another way is using simulation, or to be more specific, using **bootstrapping**. The term **bootstrapping** comes from the phrase "pulling oneself up by one's bootstraps", which is a metaphor for accomplishing an impossible task without any outside help. In this case the impossible task is estimating a population parameter (the unknown population proportion), and we'll accomplish it using data from only the given sample. Note that this notion of saying something about a population parameter using only information from an observed sample is the crux of statistical inference, it is not limited to bootstrapping. 

In essence, bootstrapping assumes that there are more of observations in the populations like the ones in the observed sample. So we "reconstruct" the population by resampling from our sample, with replacement. The bootstrapping scheme is as follows:

- **Step 1.** Take a bootstrap sample - a random sample taken **with replacement** from the original sample, of the same size as the original sample.
- **Step 2.** Calculate the bootstrap statistic - a statistic such as mean, median, proportion, slope, etc. computed on the bootstrap samples.
- **Step 3.** Repeat steps (1) and (2) many times to create a bootstrap distribution - a distribution of bootstrap statistics.
- **Step 4.** Calculate the bounds of the XX% confidence interval as the middle XX% j knof the bootstrap distribution.

Instead of coding up each of these steps, we will construct confidence intervals using the **infer** package.

Below is an overview of the functions we will use to construct this confidence interval:

Function    | Purpose
----------- | -------
`specify`   | Identify your variable of interest
`generate`  | The number of samples you want to generate
`calculate` | The sample statistic you want to do inference with, or you can also think of this as the population parameter you want to do inference for
`get_ci`    | Find the confidence interval

This code will find the 95 percent confidence interval for proportion of US adults who think climate change affects their local community.

```{r confidence interval infer}
samp %>%
  specify(response = climate_change_affects, success = "Yes") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "prop") %>%
  get_ci(level = 0.95)
```

- In `specify` we specify the `response` variable and the level of that variable we are calling a `success`.
- In `generate` we provide the number of resamples we want from the population in the `reps` argument (this should be a reasonably large number) as well as the type of resampling we want to do, which is `"bootstrap"` in the case of constructing a confidence interval.
- Then, we `calculate` the sample statistic of interest for each of these resamples, which is `prop`ortion.

Feel free to test out the rest of the arguments for these functions, since these commands will be used together to calculate confidence intervals and solve inference problems for the rest of the semester. But we will also walk you through more examples in future chapters.

To recap: even though we don't know what the full population looks like, we're 95% confident that the true proportion of US adults who think climate change affects their local community is between the two bounds reported as result of this pipeline.

## Confidence levels

## Excercise 2. In the interpretation above, we used the phrase "95% confident". What does "95% confidence" mean?

**Insert your answer here**

"95% confidence" means that you believe there is a 95% probability that the true value falls within the specified interval but there is still a small chance (5%) that it could be incorrect or due to random variation.



In this case, you have the rare luxury of knowing the true population proportion (62%) since you have data on the entire population. 

## Excercise 4. Does your confidence interval capture the true population proportion of US adults who think climate change affects their local community? If you are working on this lab in a classroom, does your neighbor's interval capture this value? 

**Insert your answer here**

Since population proportion of 0.62 falls within my confidence interval 0.4666667-0.7166667, my interval captures the true population proportion of US adults who think climate change affects their local community.


## Excercise 5.  Each student should have gotten a slightly different confidence interval. What proportion of those intervals would you expect to capture the true population mean? Why?

**Insert your answer here**

Since we constructed a 95% confidence intervals for the population proportion, then I would expect that, on average, 95% of all students will capture the true population proportion. This is because the confidence level of 95% means that, if you were to repeatedly take samples from the same population and construct confidence intervals using the same method, you would expect about 95% of those intervals to contain the true population proportion and about 5% to not contain it.


```{r shiny, echo=FALSE, eval=FALSE, results = TRUE}
# This R chunk will only run in interactive mode
store_ci <- function(i, n, reps, conf_level, success) {
  us_adults %>%
    sample_n(size = n) %>%
    specify(response = climate_change_affects, success = success) %>%
    generate(reps, type = "bootstrap") %>%
    calculate(stat = "prop") %>%
    get_ci(level = conf_level) %>%
    rename(
      x_lower = names(.)[1],
      x_upper = names(.)[2]
    )
}
library(shiny)
shinyApp(
  ui <- fluidPage(
    h4("Confidence intervals for the proportion of US adults who think 
     climate change"),

    h4(selectInput("success", "",
      choices = c(
        "is affecting their local community" = "Yes",
        "is not affecting their local community" = "No"
      ),
      selected = "Yes", width = "50%"
    )),

    # Sidebar with a slider input for number of bins
    sidebarLayout(
      sidebarPanel(
        numericInput("n_samp",
          "Sample size for a single sample from the population:",
          min = 1,
          max = 1000,
          value = 60
        ),

        hr(),

        numericInput("n_rep",
          "Number of resamples for each bootstrap confidence interval:",
          min = 1,
          max = 15000,
          value = 1000
        ),

        numericInput("conf_level",
          "Confidence level",
          min = 0.01,
          max = 0.99,
          value = 0.95,
          step = 0.05
        ),

        hr(),

        radioButtons("n_ci",
          "Number of confidence intervals:",
          choices = c(10, 25, 50, 100),
          selected = 50, inline = TRUE
        ),

        actionButton("go", "Go")
      ),

      # Show a plot of the generated distribution
      mainPanel(
        plotOutput("ci_plot")
      )
    )
  ),

  server <- function(input, output) {

    # set true p
    p <- reactive(ifelse(input$success == "Yes", 0.62, 0.38))

    # create df_ci when go button is pushed
    df_ci <- eventReactive(input$go, {
      map_dfr(1:input$n_ci, store_ci,
        n = input$n_samp,
        reps = input$n_rep, conf_level = input$conf_level,
        success = input$success
      ) %>%
        mutate(
          y_lower = 1:input$n_ci,
          y_upper = 1:input$n_ci,
          capture_p = ifelse(x_lower < p() & x_upper > p(), "Yes", "No")
        )
    })

    # plot df_ci
    output$ci_plot <- renderPlot({
      ggplot(df_ci()) +
        geom_segment(aes(x = x_lower, y = y_lower, xend = x_upper, yend = y_upper, color = capture_p)) +
        geom_point(aes(x = x_lower, y = y_lower, color = capture_p)) +
        geom_point(aes(x = x_upper, y = y_upper, color = capture_p)) +
        geom_vline(xintercept = p(), color = "darkgray") +
        labs(
          y = "", x = "Bounds of the confidence interval",
          color = "Does the interval capture the true population proportion?"
        ) +
        theme(legend.position = "bottom")
    })
  },
  options = list(height = 700)
)
```


# Excercise 6.  Given a sample size of 60, 1000 bootstrap samples for each interval, and 50 confidence intervals constructed (the default values for the above app), what proportion of your confidence intervals include the true population proportion? Is this proportion exactly equal to the confidence level? If not, explain why. Make sure to include your plot in your answer.

**Insert your answer here**

94% (47/50) of my confidence intervals include the true population proportion. This proportion is not exactly equal to the confidence level of 95%. This is because bootstrapping provides an empirical estimate of the sampling distribution, and there can be variability due to the randomness in the resampling process.
* * *

# Excercise 7. Choose a different confidence level than 95%. Would you expect a confidence interval at this level to be wider or narrower than the confidence interval you calculated at the 95% confidence level? Explain your reasoning.

**Insert your answer here**
When I chose a 99% confidence level, my confidence interval is 0.4333333 - 0.75. This is wider than the 95% confidence interval of 0.4666667 - 0.7166667. The width of a confidence interval is influenced, in part, by the chosen confidence level. If you choose a confidence level that is higher than 95%, such as a 99% confidence level, the confidence interval will be wider than the one calculated at a 95% confidence level. This is because a higher confidence level means you want to be more certain that your interval contains the true population parameter. To achieve higher confidence, you need to include a larger range of values, which results in a wider interval.

Conversely, if you choose a confidence level lower than 95%, like a 90% confidence level, the confidence interval will be narrower. A lower confidence level means you are willing to accept a higher risk of not capturing the true parameter within the interval, so you can use a narrower range of values.

In summary, confidence intervals become wider as you increase the confidence level and narrower as you decrease the confidence level.


# Excercise 8. Using code from the **infer** package and data from the one sample you have (`samp`), find a confidence interval for the proportion of US Adults who think climate change is affecting their local community with a confidence level of your choosing (other than 95%) and interpret it.

```{r}
set.seed(200)
samp %>%
specify(response = climate_change_affects, success = "Yes") %>%
generate(reps = 1000, type = "bootstrap") %>%
calculate(stat = "prop") %>%
get_ci(level = 0.99)
```

**Insert your answer here**

With 99% confidence, we estimate that the proportion of US adults who believe climate change is affecting their local community lies between 0.4333333 and 0.75. This suggests that there is a high likelihood that the true population proportion falls within this interval than that of a 95% confidence level.

# Excercise 9. Using the app, calculate 50 confidence intervals at the confidence level you chose in the previous question, and plot all intervals on one plot, and calculate the proportion of intervals that include the true population proportion. How does this percentage compare to the confidence level selected for the intervals?
    
**Insert your answer here**
Using the 99% confidence level used in the question, 100% of the intervals contain the true population proportion. This percentage is only 1% greater than the confidence level selected for the intervals. This indicates that my chosen confidence level is about appropriate for my data.


# Excercise 10. Lastly, try one more (different) confidence level. First, state how you expect the width of this interval to compare to previous ones you calculated. Then, calculate the bounds of the interval using the **infer** package and data from `samp` and interpret it. Finally, use the app to generate many intervals and calculate the proportion of intervals that are capture the true population proportion.

**Insert your answer here**
I am going to try a 90% confidence level. I expect the width of the interval to be narrower than the one for 99% confidence level. This is because a lower confidence level means I am willing to accept a higher risk of not capturing the true parameter within the interval, so I can use a narrower range of values.

# Excercise 11.  Using the app, experiment with different sample sizes and comment on how the widths of intervals change as sample size changes (increases and decreases).

**Insert your answer here**
94% of my intervals captured the true population proportion using a 90% confidence level. This is less than the 100% intervals captured for a 99% confidence interval. This is because a 90% confidence interval will capture a narrower range of the values.

# Excercise 12.  Finally, given a sample size (say, 60), how does the width of the interval change as you increase the number of bootstrap samples. **Hint:** Does changing the number of bootstap samples affect the standard error?

**Insert your answer here**
As the number of bootstrap samples increased, the precision of my estimates increased  and the confidence intervals go narrower. As I increased the number of bootstrap samples, the standard error estimate becomes more stable and accurate.
* * *